Discrete Quantum Theories and Computing

### 2019-01-25 dissertation.tex
dissertation.lyx "Chapter 1 Introduction" and Acknowledgments Updating...

I found the sentences in my Acknowledgments almost always contains
“thank”, and looks boring. To improve my Acknowledgments, I
briefly read the Acknowledgments in other theses \cite{Chih2013,Sun2015,Lee2013,Chen2016,sabry1995formal}.
\end{comment}
\begin{acknowledgments*}%
\begin{comment}
\todo{Confirm people to be acknowledged like the way being acknowledged\ldots{}
}
\end{comment}
I want to thank Prof.~Gerardo Ortiz, Prof.~Amr Sabry (\textarabic{عمرو صبري}),
and Prof.~Andrew Hanson for publishing three papers \cite{geometry2013,DQT2014,THOS2018}
with me, and enormous other things. These papers become the main part
of my thesis, especially most of Chapter~\ref{chap:DQT=000026DQC}
is based on ``Geometry of discrete quantum computing''~\cite{geometry2013}
and ``Discrete quantum theories''~\cite{DQT2014}; most of Chapter~\ref{chap:QIVPM}
is based on ``Quantum Interval-Valued Probability: Contextuality
and the Born Rule''~\cite{THOS2018}. Together with Prof.~Dylan
Thurston and Prof.~Shouhong Wang (汪守宏), they getting together to
understand my dissertation proposal and dissertation, and all the
suggestions are extremely helpful, where the proposal provides the
framework of the dissertation. Some proposed questions are solved
in Chapter~\ref{chap:QIVPM}, and some of the remaining questions
is explained in Chapter~\ref{chap:Further-Discussion}. \todo{Acknowledgments
cleaned up until here. }Also thank Prof.\ Dylan Thurston for serving
in my advisory committee and my Tier 3 committee, with other advisory
committee members passing my Computer Science qualifying exam Spring
2016. Especially, my survey of real computation in the qualifying
exam inspired me the comparison of quantum computer and analog computer
in Chapter~\ref{chap:Introduction}. Chapter~\ref{chap:Introduction}
also contains the motivation in my application for Rethinking Foundations
of Physics 2017 workshop reviewed by Prof.\ Amr Sabry~(\textarabic{عمرو صبري}).
I want to thank Prof.\ Lawrence Moss for inviting me to present our
results in the interdisciplinary logic seminar and theory seminar
in Computer Science, co-chairing my Tier 3 committee, writing an assessment
letter for me, and drawing my attention on Prof.\ Abramsky's paper~\cite{Abramsky2012},
which inspired me to merge quantum probability with discrete quantum
theories. And I thank John Gardiner for inspiring discussion on the
difficulties merging them~\cite{Gardiner2014}. These results are
improved and motivated the transition from DQT to QIVPM in Sec.~\ref{sec:Toward-IVPM}.
I also thank Prof.~Tom Lewis for helping me organize the unpublished
deterministic quantum algorithms in the term paper for SLST-T501 Academic
Writing course as presented in Sec.~\ref{subsec:Discrete-Deutsch-algorithm}.
I also thank Traci Nagle, Elizabeth ``Betsy'' Merceron, and Kexin
``Casey'' Chen (谌可心) for teaching me English in general so that
others can understand me more easily.

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick is made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
be an integer multiple of the elementary charge. A perfect square
cannot be drawn physically. Even we don't draw it, but merely ask
whether it exists in the space. This question is still beyond our
ability to answer because whether space itself is ultimately discrete
or continuous is still an unsolved question. Even if a perfect square
exists physically, we may not able to decide whether the ``square''
we discussed is really perfect or almost perfect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation closes to
zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by a realistic analog computer. Since two branches
of computations may not have any relations, the follow-up computational
paths may be significantly different, and this difference is unlikely
to be compensated by error analysis techniques easily like the butterfly
effect. Therefore, we might not be able to utilize the realistic analog
computers to acquire the computational power predicted by this kind
of theoretical models.%
\begin{comment}
\todo{Maybe extended to consider probabilistic model? }
\end{comment}

@@

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{YcYY}
\toprule 
\addlinespace
Physical quantity & Discrete & Discrete with extreme small units & Continuous, or no evidence to support they are discrete\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational device & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integer & Real number & Real number\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How does the theorical model predict the behavior of the physical device? & 
Reliably & 
Not reliably, but after we have better technology to manipulate the minimum units, we can more reliably predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be able compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum model &  & Quantum theories and computing over finite fields & Quantum interval-valued probability measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and two
possibilities for classical quantities inspires two types of quantum
models. Like its classical counterparts, quantum circuit model, the
most widely used model of quantum computing, manipulates quantities
which are not discrete apparently: the probability amplitudes of quantum
states, which are assumed in the field of complex numbers in the ``conventional
quantum theory'' (CQT)\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}. If the probability amplitudes are assumed to be ultimately discrete,
we choose to replace the field of complex numbers by discrete finite
fields for two reasons. On one hand, CQT is built upon the operations
among the probability amplitudes, and the probability amplitudes have
the required operations since the complex numbers form a field. On
the other hand, based on finite fields, we will define the fraction-like
cardinal probability which has extremely small units when the size
of the field is extremely large. Alternatively, even if the probability
amplitudes might never be found quantized, considering the precision
of quantities could still result in a better model. However, comparing
to the precision of probability amplitudes, it is easier to consider
the precision of their inducing probabilities because the idea of
``imprecise probabilities'' has well-studied classically \cite{Shafer1976,GilboaSchmeidler1994,Marinacci1999,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
\begin{comment}
\todo{Make sure whether realistic quantum computers could branch
their computation based on their probability amplitudes or measured
probabilities? If they can only branch on their measured probabilities,
it will provide another reason why we want to study the precision
of the inducing probabilities\ldots{} }
\end{comment}
This idea will be extended to the quantum domain as quantum interval-valued
probability measures (QIVPM).

>>>>

I found the sentences in my Acknowledgments almost always contains
“thank”, and looks boring. To improve my Acknowledgments, I
briefly read the Acknowledgments in other theses \cite{Chih2013,Sun2015,Lee2013,Chen2016,sabry1995formal}.
\end{comment}
\begin{acknowledgments*}%
\begin{comment}
\todo{Confirm people to be acknowledged like the way being acknowledged\ldots{}
}
\end{comment}
I want to thank Prof.~Gerardo Ortiz, Prof.~Amr Sabry (\textarabic{عمرو صبري}),
and Prof.~Andrew Hanson for publishing three papers \cite{geometry2013,DQT2014,THOS2018}
with me, and enormous other things. These papers become the main part
of my thesis, especially most of Chapter~\ref{chap:Conventional-Quantum-Theory},
\ref{chap:Unrestricted-Finite-Fields}, and~\ref{chap:DQT=000026DQC}
is based on ``Geometry of discrete quantum computing''~\cite{geometry2013}
and ``Discrete quantum theories''~\cite{DQT2014}; most of Chapter~\ref{chap:QIVPM}
is based on ``Quantum Interval-Valued Probability: Contextuality
and the Born Rule''~\cite{THOS2018}. Together with Prof.~Dylan
Thurston and Prof.~Shouhong Wang (汪守宏), you provide helpful suggestions
for my dissertation proposal and dissertation itself, where the proposal
provides the framework of the dissertation. Some proposed questions
are solved in Chapter~\ref{chap:QIVPM}, and some of the remaining
questions are explained in Chapter~\ref{chap:Further-Discussion}.
While being completely rewritten, the idea of Chapter~\ref{chap:Introduction}
comes from my application for Rethinking Foundations of Physics 2017
workshop reviewed by Prof.\ Amr Sabry~(\textarabic{عمرو صبري})
and my survey of real computation in my Computer Science qualifying
exam with my advisory committee, Prof.~Amr Sabry(\textarabic{عمرو صبري}),
Prof.~Gerardo Ortiz, and Prof.~Andrew Hanson, and Prof.\ Dylan
Thurston. My study for Tier 3 exam, which is part of the qualifying
exam for Mathematics, is also helpful when I start to export the work
of quantum computing, and \todo{Acknowledgments cleaned up until
here. }Also thank Prof.\ Dylan Thurston for serving in my advisory
committee and my Tier 3 committee, with other advisory committee members
passing my Computer Science qualifying exam Spring 2016. Especially,
my survey of real computation in the qualifying exam inspired me the
comparison of quantum computer and analog computer in Chapter~\ref{chap:Introduction}.
Chapter~\ref{chap:Introduction} also contains the motivation in
my application for Rethinking Foundations of Physics 2017 workshop
reviewed by Prof.\ Amr Sabry~(\textarabic{عمرو صبري}). I
want to thank Prof.\ Lawrence Moss for inviting me to present our
results in the interdisciplinary logic seminar and theory seminar
in Computer Science, co-chairing my Tier 3 committee, writing an assessment
letter for me, and drawing my attention on Prof.\ Abramsky's paper~\cite{Abramsky2012},
which inspired me to merge quantum probability with discrete quantum
theories. And I thank John Gardiner for inspiring discussion on the
difficulties merging them~\cite{Gardiner2014}. These results are
improved and motivated the transition from DQT to QIVPM in Sec.~\ref{sec:Toward-IVPM}.
I also thank Prof.~Tom Lewis for helping me organize the unpublished
deterministic quantum algorithms in the term paper for SLST-T501 Academic
Writing course as presented in Sec.~\ref{subsec:Discrete-Deutsch-algorithm}.
I also thank Traci Nagle, Elizabeth ``Betsy'' Merceron, and Kexin
``Casey'' Chen (谌可心) for teaching me English in general so that
others can understand me more easily.

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick is made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
be an integer multiple of the elementary charge. A perfect square
cannot be drawn physically. Even we don't draw it, but merely ask
whether it exists in the space. This question is still beyond our
ability to answer because whether space itself is ultimately discrete
or continuous is still an unsolved question. Even if a perfect square
exists physically, we may not able to decide whether the ``square''
we discussed is really perfect or almost perfect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation closes to
zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by a realistic analog computer. Since two branches
of computations may not have any relations, the follow-up computational
paths may be significantly different, and this difference is unlikely
to be compensated by error analysis techniques easily like the butterfly
effect. Therefore, we might not be able to utilize a realistic analog
computer to acquire the computational power predicted by this kind
of theoretical models.%
\begin{comment}
\todo{Maybe extended to consider probabilistic model? }
\end{comment}

@@

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{YcYY}
\toprule 
\addlinespace
Physical quantity & Discrete & Discrete with extreme small units & Continuous, or no evidence to support they are discrete\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational device & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integer & Real number & Real number\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How does the theorical model predict the behavior of the physical device? & 
Reliably & 
Not reliably, but after we have better technology to manipulate the minimum units, we can more reliably predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be able compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum model &  & Quantum theories and computing over finite fields & Quantum interval-valued probability measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and two
possibilities for classical quantities inspires two types of quantum
models. Like its classical counterparts, quantum circuit model, the
most widely used model of quantum computing, manipulates quantities
which are not discrete apparently: the probability amplitudes of quantum
states, which are assumed in the field of complex numbers in the ``conventional
quantum theory'' (CQT)\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}. If the probability amplitudes are assumed to be ultimately discrete,
we choose to replace the field of complex numbers by discrete finite
fields for two reasons. On one hand, CQT is built upon the operations
among the probability amplitudes, and the probability amplitudes have
the required operations since the complex numbers form a field. On
the other hand, based on finite fields, we will define the fraction-like
cardinal probability which has extremely small units when the size
of the field is extremely large. Alternatively, even if the probability
amplitudes might never be found quantized, considering the precision
of quantities could still result in a better model. However, comparing
to the precision of probability amplitudes, it is easier to consider
the precision of their inducing probabilities because the idea of
``imprecise probabilities'' has well-studied classically \cite{Shafer1976,GilboaSchmeidler1994,Marinacci1999,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
\begin{comment}
\todo{Make sure whether realistic quantum computers could branch
their computation based on their probability amplitudes or measured
probabilities? If they can only branch on their measured probabilities,
it will provide another reason why we want to study the precision
of the inducing probabilities\ldots{} }
\end{comment}
This idea will be extended to quantum domain as quantum interval-valued
probability measures (QIVPM).
